---
title: "Assignment #1 (HW #1)"
subtitle: "Part 1: Annotating `{ggplot2}` code"
description: "Assigned Mon 01/06/2025 | Due Sat 01/18/2025"
author: "Your Name"
toc: true
editor_options: 
  chunk_output_type: console
---

# Part 1a. Annotate 

## I. Setup

```{r}
#| eval: false
#| echo: true

#..........................load packages.........................
library(colorspace)
library(geofacet) 
library(ggtext) 
library(glue) 
library(grid)
library(magick)
library(patchwork) 
library(scales) 
library(showtext) 
library(tidyverse) 

#.............read in data from GitHub (tidytuesday).............
ufo_sightings <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/ufo_sightings.csv')
places <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/places.csv')

#.........create color palette and named color variables.........
alien <- c("#101319", "#28ee85")
bg <- alien[1]
accent <- alien[2]

#........................import UFO image........................
ufo_image <- magick::image_read(path = here::here("images", "ufo.png")) 

#......................import Google fonts.......................
sysfonts::font_add_google(name = "Orbitron", family = "orb")
sysfonts::font_add_google(name = "Barlow", family = "bar")

#....................import FontAwesome fonts....................
sysfonts::font_add(family = "fa-brands", regular = here::here("fonts", "Font Awesome 6 Brands-Regular-400.otf"))
sysfonts::font_add(family = "fa-solid", regular = here::here("fonts", "Font Awesome 6 Free-Solid-900.otf"))

#........use {showtext} to render text for future devices........
showtext::showtext_auto(enable = TRUE)
```

## II. Data wrangling

### i. Create df of population size by US state

```{r}
#| eval: false
#| echo: true

df_pop <- places |>
  
  # filter the `country_code` column for just US observations ----
  filter(country_code == "US") |>
  
  # update incorrect state code, "Fl", to "FL" ----
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  
  # calculate the total population size by state and add as a column called `pop` ----
  group_by(state) |>
  summarise(pop = sum(population)) |>
  ungroup()
```

### ii. Create df of sightings and population by state

```{r}
#| eval: false
#| echo: true

df_us <- ufo_sightings |>
  
  # ANNOTATE ----
  filter(country_code == "US") |>
  
  # update incorrect state code, "Fl", to "FL" ----
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  
  # ANNOTATE ----
  count(state) |>
  
  # ANNOTATE ----
  left_join(df_pop, by = "state") |>
  
  # rename column `n` (for clarity) ----
  rename(num_obs = n) |> 
  
  # add new columns ----
  mutate(
    
    # ANNOTATE ----
    num_obs_per10k = num_obs / pop * 10000,
    
    # ANNOTATE (be sure to describe what the calculated values in this new column represent) ----
    opacity_val = num_obs_per10k / max(num_obs_per10k)
    
    )
```

### iii. Create df of most commonly sighted UFO shapes

```{r}
#| eval: false
#| echo: true

df_shape <- ufo_sightings |>
  
  # ANNOTATE ----
  filter(!shape %in% c("unknown", "other")) |>
  
  # ANNOTATE ----
  count(shape) |>
  
  # rename column `n` (for clarity) ----
  rename(total_sightings = n) |> 
  
  # ANNOTATE ----
  arrange(desc(total_sightings)) |>
  
  # ANNOTATE ----
  slice_head(n = 10) |>
  
  # add new columns ----
  mutate(
    
    # ANNOTATE ----
    shape = fct_reorder(.f = shape, 
                        .x = total_sightings), 
    
    # ANNOTATE (be sure to describe what the calculated values in this new column represent) ----
    opacity_val = scales::rescale(x = total_sightings, 
                                  to = c(0.3, 1))

    )
```

### iv. Create df of sightings by day / time

```{r}
#| eval: false
#| echo: true

df_day_hour <- ufo_sightings |>
  
  # create three new cols with day number, hour, and day abbreviation, extracted from the `reported_date_time` column ----
  mutate(
    day = wday(reported_date_time), 
    hour = hour(reported_date_time), 
    wday = wday(reported_date_time, label = TRUE) 
  ) |>
  
  # ANNOTATE ----
  count(day, wday, hour) |>
  
  # rename column `n` (for clarity) ----
  rename(total_daily_obs = n) |> 
  
  # add new columns ----
  mutate(
    
    # ANNOTATE (be sure to describe what the calculated values in this new column represent) ----
    opacity_val = total_daily_obs / max(total_daily_obs),
    
    # ANNOTATE ----
    hour_lab = case_when(
      hour == 0 ~ "12am",
      hour <= 12 ~ paste0(hour, "am"),
      hour == 12 ~ "12pm",
      TRUE ~ paste0(hour - 12, "pm")) 
    
    )
```

## III. Prepare text elements

```{r}
#| eval: false
#| echo: true

#........extract & format quotes from `ufo_sightings` df.........
quotes <- paste0('"...', str_to_sentence(ufo_sightings$summary[c(47816, 6795, 93833)]), '..."')

#....create caption (which includes fontawesome icons & text)....
original <- glue("Original visualization by Dan Oehm:")
dan_github <- glue("<span style='font-family:fa-brands;'>&#xf09b;</span> doehm/tidytues")
new <- glue("Updated version by Sam Shanny-Csik for EDS 240:")
link <- glue("<span style='font-family:fa-solid;'>&#xf0c1;</span> eds-240-data-viz.github.io")
space <- glue("<span style='color:{bg};'>. .</span>")
caption <- glue("{original}{space}{dan_github}
                <br><br>
                {new}{space}{link}")
```

## IV. Build plots

### i. Create bar plot of commonly reported UFO shapes

```{r}
#| eval: false
#| echo: true

plot_shape <- ggplot(data = df_shape) +
  
  # ANNOTATE (be sure to comment on the type of geom & describe aesthetic mappings) ----
  geom_col(aes(x = total_sightings, y = shape, alpha = opacity_val), 
           fill = accent) +
  
  # add shape labels (in Title case) to the top left corner of bars ----
  geom_text(aes(x = 200, y = shape, label = str_to_title(shape)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 14, 
            hjust = 0,
            nudge_y = 0.2) +
  
  # ANNOTATE (be sure to describe aesthetics mappings, similar to the geom_text() layer, directly above) ----
  geom_text(aes(x = total_sightings-200, y = shape, label = scales::comma(total_sightings)),
            family = "orb",
            fontface = "bold",
            color = bg,
            size = 10,
            hjust = 1,
            nudge_y = -0.2) +
  
  # ANNOTATE ----
  scale_x_continuous(expand = c(0, NA)) +

  # ANNOTATE ----
  labs(subtitle = "10 most commonly reported shapes") +

  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
    # ANNOTATE ---
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 0,  
                                 margin = margin(b = 10)),
    
    # ANNOTATE ----
    legend.position = "none"
    
  )
```

### ii. Create US map with sightings by state 

**HINT:** Consider temporarily commenting out / rearranging the `geom_*()` layers to better understand how this plot is constructed

```{r}
#| eval: false
#| echo: true

plot_us <-  ggplot(df_us) +

  # ANNOTATE ----
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1, alpha = opacity_val), 
            fill = accent) +
  
  # ANNOTATE ----
  geom_text(aes(x = 0.5, y = 0.7, label = state), 
            family = "orb", 
            fontface = "bold",
            size = 9, 
            color = bg) +
  
  # ANNOTATE ----
  geom_text(aes(x = 0.5, y = 0.3, label = round(num_obs_per10k, 1)), 
            family = "orb", 
            fontface = "bold",
            size = 8, 
            color = bg) +  

  # ANNOTATE ----
  geofacet::facet_geo(~state) +

  # ANNOTATE ----
  coord_fixed(ratio = 1) +
  
  # ANNOTATE ----
  labs(subtitle = "Sightings per 10k population") +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
    # ANNOTATE ----
    strip.text = element_blank(),
    
    # ANNOTATE ----
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 1,  
                                 margin = margin(b = 10)),
    
    # ANNOTATE ----
    legend.position = "none"
    
  )
```

### iii. Create tile plot of sightings by day / time

```{r}
#| eval: false
#| echo: true

plot_day <- ggplot(data = df_day_hour) +
  
  # ANNOTATE (be sure to comment on the type of geom & describe aesthetic mappings) ---- 
  geom_tile(aes(x = hour, y = day, alpha = opacity_val), 
            fill = accent, 
            height = 0.9, 
            width = 0.9) +
  
  # ANNOTATE (be sure to describe aesthetics mappings) ----
  geom_text(aes(x = hour, y = 9, label = hour_lab), 
            family = "orb",
            color = accent, 
            size = 10) +
  
  # ANNOTATE (be sure to describe aesthetics mappings) ----
  geom_text(aes(x = 0, y = day, label = str_sub(string = wday, start = 1, end = 1)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 8) + 
  
  # ANNOTATE (also comment on ~why~ we're doing this) ----
  ylim(-5, 9) +
  xlim(NA, 23.55) +
  
  # ANNOTATE ----
  coord_polar() +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
    # ANNOTATE ----
    plot.background = element_rect(fill = bg, color = bg),
    
    # ANNOTATE ----
    legend.position = "none"
    
  )
```

### iv. Create quotes as individual plot objects

A comment from Dan Oehm's original code: "A bit clunky but the path of least resistance."

**Note:** You only need to annotate `quote1` (`quote2` & `quote3` are constructed nearly identically).

```{r}
#| eval: false
#| echo: true

#............................quote 1.............................
quote1 <- ggplot() +
  
  # ANNOTATE ----
  annotate(geom ="text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[1], width = 40),
           family = "bar", 
           fontface = "italic", 
           color = accent, 
           size = 16, 
           hjust = 0, 
           lineheight = 0.4) +
  
  # ANNOTATE ----
  xlim(0, 1) +
  ylim(0, 1) +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # ANNOTATE (be sure to describe why we must set `clip = "off"`) ----
  coord_cartesian(clip = "off")

#............................quote 2.............................
quote2 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[2], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")

#............................quote 3.............................
quote3 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[3], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")
```

### v. Turn UFO image into a ggplot object

**Note:** Grob stands for **gr**aphical **ob**ject. Each visual element rendered in a a ggplot (e.g. lines, points, axes, entire panels, even images) is represented as a grob. Grobs can be manipulated individually to fully customize plots. 

```{r}
#| eval: false
#| echo: true
plot_ufo <- ggplot() +
  
  # convert magick image to a grob and add as an annotation ----
  annotation_custom(grid::rasterGrob(ufo_image)) +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
    # ANNOTATE ----
    plot.background = element_rect(fill = bg, color = bg)
    
  )
```

### vi. Create infographic base plot (the plot space that all other plots will be added to)

```{r}
#| eval: false
#| echo: true

plot_base <- ggplot() +
  
  # ANNOTATE ----
  labs(
    title = "UFO Sightings",
    subtitle = "Summary of over 88k reported sightings across the US",
    caption = caption
    ) +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
   # ANNOTATE ----
    text = element_text(family = "orb", 
                        size = 48, 
                        lineheight = 0.3, 
                        color = accent),
    
    # ANNOTATE ----
    plot.background = element_rect(fill = bg, 
                                   color = bg),
    
    # ANNOTATE ----
    plot.title = element_text(size = 128, 
                              face = "bold", 
                              hjust = 0.5, 
                              margin = margin(b = 10)),
    
    # ANNOTATE ----
    plot.subtitle = element_text(family = "bar", 
                                 hjust = 0.5, 
                                 margin = margin(b = 20)),
    
    # format plot caption using {ggtext} to render fontawesome icons ----
    plot.caption = ggtext::element_markdown(family = "bar",
                                            face = "italic",
                                            color = colorspace::darken(accent, 0.25),
                                            hjust = 0.5,
                                            margin = margin(t = 20)),

    # ANNOTATE ----
    plot.margin = margin(b = 20, t = 50, r = 50, l = 50)
    
  )
```

## V. Stitch it all together and save

```{r}
#| eval: false
#| echo: true

#.....................stitch it all together.....................
plot_final <- plot_base +
  
  # ANNOTATE ----
  inset_element(plot_shape, left = 0, right = 1, top = 1, bottom = 0.66) +
  inset_element(plot_us, left = 0.42, right = 1, top = 0.74, bottom = 0.33) +
  inset_element(plot_day, left = 0, right = 0.66, top = 0.4, bottom = 0) +
  inset_element(quote1, left = 0.5, right = 1, top = 0.8, bottom = 0.72) +
  inset_element(quote2, left = 0, right = 1, top = 0.52, bottom = 0.4) +
  inset_element(quote3, left = 0.7, right = 1, top = 0.2, bottom = 0) +
  inset_element(plot_ufo, left = 0.25, right = 0.41, top = 0.23, bottom = 0.17) + 
  
  # ANNOTATE (it's helpful to comment this out, then save the plot using `ggsave()` to understand why this part is necessary) ----
  plot_annotation(
    theme = theme(
      plot.background = element_rect(fill = bg,
                                     color = bg)
    )
  ) 

#.......................save as a png file.......................
ggsave(plot = plot_final, 
       filename = here::here("outputs", "ufo_sightings_infographic.png"), 
       height = 16, 
       width = 10)
```

# Part 1b: Answer questions 

1. **Consider the order of `geom_*()` layers in the US map plot (`plot_us`). Why did the original creator order the layers in this way?**

    - Answer here
    
2. **During week 2, we discuss [Choosing the right graphic form](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture2.1-choosing-graphic-forms-slides.html#/title-slide). Refer to this lecture when answering the sub-questions, below:**

    a. **What "perceptual tasks" (from Cleveland & McGill's heirarchy) must the viewer perform to extract information from these visualizations?**
        
        - Answer here
    
    b. **What task(s) do you think the author wanted to enable or message(s) he wanted to convey with these visualizations (see lecture 2.1, slide 16 for examples)? Be sure to note at least one task / message for each of the three data viz.**
    
        - Answer here
    
    c. **Name at least one caveat to the "hierarchy of perceptual tasks" that the author employed to achieve a goal(s) you noted in question b?**
    
       - Answer here

3. **Describe two elements of this piece that you find visually-pleasing / easy to understand / intuitive. Why?** 

    - Answer here 

4. **Describe two elements of this piece that you feel could be better presented in a different way. Why?** 

    - Answer here 
    
5. **Describe two new things that you learned by interpreting / annotating this code. These could be packages, functions, or even code organizational approaches that you hadn't previously known about or considered.**

    - Answer here
    
