---
title: "HW #1"
subtitle: "Interpreting `{ggplot2}` code"
description: "Assigned Wed 01/07/2026 | Due Wed 01/14/2026"
author: "Your Name"
toc: true
code-line-numbers: true
editor_options: 
  chunk_output_type: console
---

## I. Setup

```{r}
#| eval: false
#| echo: true

library(colorspace)
library(geofacet) 
library(ggtext) 
library(glue) 
library(grid)
library(magick)
library(patchwork) 
library(scales) 
library(showtext) 
library(tidyverse) 

ufo_sightings <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/ufo_sightings.csv')
places <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/places.csv')

alien <- c("#101319", "#28ee85")
bg <- alien[1]
accent <- alien[2]

ufo_image <- magick::image_read(path = here::here("images", "ufo.png")) 

sysfonts::font_add_google(name = "Orbitron", family = "orb")
sysfonts::font_add_google(name = "Barlow", family = "bar")
sysfonts::font_add(family = "fa-brands", regular = here::here("fonts", "Font Awesome 6 Brands-Regular-400.otf"))
sysfonts::font_add(family = "fa-solid", regular = here::here("fonts", "Font Awesome 6 Free-Solid-900.otf"))

showtext::showtext_auto(enable = TRUE)
```

1. **What is the author defining in lines 15-17? Where else in the code do these defined variables show up? What advantages are there to defining these values here, as variables, rather than defining the values directly throughout the script?**

    - Your answer here

2. **In your own words, explain what the function, `font_add_google()`, does. What's the difference between the two arguments, `name` and `family`?**

    - Your answer here

## II. Data wrangling

### i. Create df of population size by US state

```{r}
#| eval: false
#| echo: true

df_pop <- places |>
  filter(country_code == "US") |>
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  group_by(state) |>
  summarise(pop = sum(population)) |>
  ungroup()
```

### ii. Create df of sightings and population by state

```{r}
#| eval: false
#| echo: true

df_us <- ufo_sightings |>
  filter(country_code == "US") |>
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  count(state) |>
  left_join(df_pop, by = "state") |>
  rename(num_obs = n) |> 
  mutate(
    num_obs_per10k = num_obs / pop * 10000,
    opacity_val = num_obs_per10k / max(num_obs_per10k)
    )
```

### iii. Create df of most commonly sighted UFO shapes

```{r}
#| eval: false
#| echo: true

df_shape <- ufo_sightings |>
  filter(!shape %in% c("unknown", "other")) |>
  count(shape) |>
  rename(total_sightings = n) |> 
  arrange(desc(total_sightings)) |>
  slice_head(n = 10) |>
  mutate(
    shape = fct_reorder(.f = shape, 
                        .x = total_sightings), 
    opacity_val = scales::rescale(x = total_sightings, 
                                  to = c(0.3, 1))
    )
```

### iv. Create df of sightings by day / time

```{r}
#| eval: false
#| echo: true

df_day_hour <- ufo_sightings |>
  mutate(
    day = wday(reported_date_time), 
    hour = hour(reported_date_time), 
    wday = wday(reported_date_time, label = TRUE) 
  ) |>
  count(day, wday, hour) |>
  rename(total_daily_obs = n) |> 
  mutate(
    opacity_val = total_daily_obs / max(total_daily_obs),
    hour_lab = case_when(
      hour == 0 ~ "12am",
      hour <= 12 ~ paste0(hour, "am"),
      hour == 12 ~ "12pm",
      TRUE ~ paste0(hour - 12, "pm")) 
    )
```

## III. Prepare text elements

```{r}
#| eval: false
#| echo: true

quotes <- paste0('"...', str_to_sentence(ufo_sightings$summary[c(47816, 6795, 93833)]), '..."')

original <- glue("Original visualization by Dan Oehm:")
dan_github <- glue("<span style='font-family:fa-brands;'>&#xf09b;</span> doehm/tidytues")
new <- glue("Updated version by Sam Shanny-Csik for EDS 240:")
link <- glue("<span style='font-family:fa-solid;'>&#xf0c1;</span> eds-240-data-viz.github.io")
space <- glue("<span style='color:{bg};'>. .</span>")
caption <- glue("{original}{space}{dan_github}
                <br><br>
                {new}{space}{link}")
```

## IV. Build plots

### i. Create bar plot of commonly reported UFO shapes

```{r}
#| eval: false
#| echo: true

plot_shape <- ggplot(data = df_shape) +
  geom_col(aes(x = total_sightings, y = shape, alpha = opacity_val), 
           fill = accent) +
  geom_text(aes(x = 200, y = shape, label = str_to_title(shape)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 14, 
            hjust = 0,
            nudge_y = 0.2) +
  geom_text(aes(x = total_sightings-200, y = shape, label = scales::comma(total_sightings)),
            family = "orb",
            fontface = "bold",
            color = bg,
            size = 10,
            hjust = 1,
            nudge_y = -0.2) +
  scale_x_continuous(expand = c(0, NA)) +
  labs(subtitle = "10 most commonly reported shapes") +
  theme_void() +
  theme(
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 0,  
                                 margin = margin(b = 10)),
    legend.position = "none"  
  )
```

### ii. Create US map with sightings by state 

**HINT:** Consider temporarily commenting out / rearranging the `geom_*()` layers to better understand how this plot is constructed

```{r}
#| eval: false
#| echo: true

plot_us <-  ggplot(df_us) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1, alpha = opacity_val), 
            fill = accent) +
  geom_text(aes(x = 0.5, y = 0.7, label = state), 
            family = "orb", 
            fontface = "bold",
            size = 9, 
            color = bg) +
  geom_text(aes(x = 0.5, y = 0.3, label = round(num_obs_per10k, 1)), 
            family = "orb", 
            fontface = "bold",
            size = 8, 
            color = bg) +  
  geofacet::facet_geo(~state) +
  coord_fixed(ratio = 1) +
  labs(subtitle = "Sightings per 10k population") +
  theme_void() +
  theme(
    strip.text = element_blank(),
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 1,  
                                 margin = margin(b = 10)),
    legend.position = "none"  
  )
```

- [ ] **tbd.** Consider the order of `geom_*()` layers in the the above plot (`plot_us`). Why did the author order the layers in this way?

### iii. Create tile plot of sightings by day / time

```{r}
#| eval: false
#| echo: true

plot_day <- ggplot(data = df_day_hour) +
  geom_tile(aes(x = hour, y = day, alpha = opacity_val), 
            fill = accent, 
            height = 0.9, 
            width = 0.9) +
  geom_text(aes(x = hour, y = 9, label = hour_lab), 
            family = "orb",
            color = accent, 
            size = 10) +
  geom_text(aes(x = 0, y = day, label = str_sub(string = wday, start = 1, end = 1)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 8) + 
  ylim(-5, 9) +
  xlim(NA, 23.55) +
  coord_polar() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = bg, color = bg),
    legend.position = "none"  
  )
```

### iv. Create quotes as individual plot objects

A comment from Dan Oehm's original code: "A bit clunky but the path of least resistance."

**Note:** You only need to annotate `quote1` (`quote2` & `quote3` are constructed nearly identically).

```{r}
#| eval: false
#| echo: true

quote1 <- ggplot() +
  annotate(geom ="text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[1], width = 40),
           family = "bar", 
           fontface = "italic", 
           color = accent, 
           size = 16, 
           hjust = 0, 
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")

quote2 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[2], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")

quote3 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[3], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")
```

- [ ] What is the purpose of, `coord_cartesian(clip = "off")`?

### v. Turn UFO image into a ggplot object

**Note:** Grob stands for **gr**aphical **ob**ject. Each visual element rendered in a a ggplot (e.g. lines, points, axes, entire panels, even images) is represented as a grob. Grobs can be manipulated individually to fully customize plots. 

```{r}
#| eval: false
#| echo: true
plot_ufo <- ggplot() +
  annotation_custom(grid::rasterGrob(ufo_image)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = bg, color = bg)
  )
```

### vi. Create infographic base plot (the plot space that all other plots will be added to)

```{r}
#| eval: false
#| echo: true

plot_base <- ggplot() +
  labs(
    title = "UFO Sightings",
    subtitle = "Summary of over 88k reported sightings across the US",
    caption = caption
    ) +
  theme_void() +
  theme(
    text = element_text(family = "orb", 
                        size = 48, 
                        lineheight = 0.3, 
                        color = accent),
    plot.background = element_rect(fill = bg, 
                                   color = bg),
    plot.title = element_text(size = 128, 
                              face = "bold", 
                              hjust = 0.5, 
                              margin = margin(b = 10)),
    plot.subtitle = element_text(family = "bar", 
                                 hjust = 0.5, 
                                 margin = margin(b = 20)),
    plot.caption = ggtext::element_markdown(family = "bar",
                                            face = "italic",
                                            color = colorspace::darken(accent, 0.25),
                                            hjust = 0.5,
                                            margin = margin(t = 20)),
    plot.margin = margin(b = 20, t = 50, r = 50, l = 50)
    
  )
```

## V. Stitch it all together and save

```{r}
#| eval: false
#| echo: true

plot_final <- plot_base +
  inset_element(plot_shape, left = 0, right = 1, top = 1, bottom = 0.66) +
  inset_element(plot_us, left = 0.42, right = 1, top = 0.74, bottom = 0.33) +
  inset_element(plot_day, left = 0, right = 0.66, top = 0.4, bottom = 0) +
  inset_element(quote1, left = 0.5, right = 1, top = 0.8, bottom = 0.72) +
  inset_element(quote2, left = 0, right = 1, top = 0.52, bottom = 0.4) +
  inset_element(quote3, left = 0.7, right = 1, top = 0.2, bottom = 0) +
  inset_element(plot_ufo, left = 0.25, right = 0.41, top = 0.23, bottom = 0.17) + 
  plot_annotation(
    theme = theme(
      plot.background = element_rect(fill = bg,
                                     color = bg)
    )
  ) 

ggsave(plot = plot_final, 
       filename = here::here("outputs", "ufo_sightings_infographic.png"), 
       height = 16, 
       width = 10)
```

### Answer some final reflective questions 

1. **During week 2, we discuss [Choosing the right graphic form](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture2.1-choosing-graphic-forms-slides.html#/title-slide). Refer to this lecture when answering the sub-questions, below:**

    a. **What "perceptual tasks" (from Cleveland & McGill's heirarchy) must the viewer perform to extract information from these visualizations?**
        
        - Answer here
    
    b. **What task(s) do you think the author wanted to enable or message(s) he wanted to convey with these visualizations (see lecture 2.1, slide 16 for examples)? Be sure to note at least one task / message for each of the three data viz.**
    
        - Answer here.
    
    c. **Name at least one caveat to the "hierarchy of perceptual tasks" that the author employed to achieve a goal(s) you noted in question b?**
    
       - Answer here

2. **Describe two elements of this piece that you find visually-pleasing / easy to understand / intuitive. Why?** 

    - Answer here 

3. **Describe two elements of this piece that you feel could be better presented in a different way. Why?** 

    - Answer here 
    
4. **Describe two new things that you learned by interpreting / annotating this code. These could be packages, functions, or even code organizational approaches that you hadn't previously known about or considered.**

    - Answer here
    
    